Designing a high-performance boundary element library with OpenCL and Numba

### 2021-05-10 bem.tex
Revisions done.

Once a triangulation is given we need to define the necessary data structures for the discretization. Bempp-cl uses a Galerkin discretization: we introduce set of basis functions \(\phi_1\) to \(\phi_N\), and define the \textbf{trial space} as the span of these functions. We then approximate the solution \(\phi\) of \cref{eq:bnd_integral} by $\phi_h=\sum_{j=1}^N \dvec{x}_j\phi_j$, where $\dvec{x}$ is a vector of coefficients. In the most simple case, we can define the function \(\phi_j\) to be equal to 1 on the triangle $\tau_j$ and 0 everywhere else. Other spaces are commonly defined to be piecewise polynomials on each triangle.

To discretize \cref{eq:bnd_integral}, we define a \textbf{test (or dual) space} in terms of a basis \(\psi_1\) to \(\psi_N\). The discrete representation of the above problem then takes the form
$$
\dmat{A}\dvec{x} = \dvec{b}
$$
with
\begin{align*}
\dmat{A}_{ij} &= \int_{\Gamma}\psi_i(\bx)\int_{\Gamma}g(\bx, \by)\phi_j(\bx)\ds[\by]\ds[\bx]\\
\dvec{b}_i &= \int_{\Gamma}\psi_i(\bx)f(\bx)\ds[\bx].
\end{align*}
In the case of piecewise constant trial and test functions, the definition of $\dmat{A}_{ij}$ simplifies to $\dmat{A}_{ij} = \int_{\Gamma}\int_{\Gamma}g(\bx, \by)\ds[\by]\ds[\bx]$.

@@

Once the grid, the space objects, and the operator(s) are defined, the main computational step is performed, namely the \textbf{computation of the discrete matrix entries} $\dmat{A}_{ij}$. For pairs of triangles $\tau_i$ and $\tau_j$ that do not share a joint edge or vertex this is done through a simple numerical quadrature rule that approximates $\dmat{A}_{ij}\approx \sum_{\ell}\sum_{q} g(\bx_\ell, \by_q)\psi_i(\bx_\ell)\phi_j(\by_q)\omega_\ell\omega_q$, where the $\bx_\ell$ and $\by_q$ are quadrature points in the corresponding triangles $\tau_i$ and $\tau_j$, and the values $\omega_\ell$ and $\omega_q$ are the quadrature weights. In the case that two triangles share a joint vertex/edge or the triangles $\tau_i$ and $\tau_j$ are identical (see \cref{fig:triangles}), corresponding singular quadrature rules are used that are based on singularity-removing coordinate transformations.

@@

All of these steps are accelerated through the use of either Numba or OpenCL. In the following section we provide a high-level overview of the library and how these acceleration techniques are deployed before we deep dive into the design of the computational kernels.

>>>>

Once a triangulation is given we need to define the necessary data structures for the discretization. Bempp-cl uses a Galerkin discretization: we introduce a set of basis functions \(\phi_1\) to \(\phi_N\), and define the \textbf{trial space} as the span of these functions. We then approximate the solution \(\phi\) of \cref{eq:bnd_integral} by $\phi_h=\sum_{j=1}^N \dvec{x}_j\phi_j$, where $\dvec{x}$ is a vector of coefficients. In the most simple case, we can define the function \(\phi_j\) to be equal to 1 on the triangle $\tau_j$ and 0 everywhere else. Other spaces are commonly defined to be piecewise polynomials on each triangle.

To discretize \cref{eq:bnd_integral}, we define a \textbf{test (or dual) space} in terms of a basis \(\psi_1\) to \(\psi_N\). The discrete representation of the above problem then takes the form
$$
\dmat{A}\dvec{x} = \dvec{b}
$$
with
\begin{align*}
\dmat{A}_{ij} &= \int_{\Gamma}\psi_i(\bx)\int_{\Gamma}g(\bx, \by)\phi_j(\bx)\ds[\by]\ds[\bx]\\
\dvec{b}_i &= \int_{\Gamma}\psi_i(\bx)f(\bx)\ds[\bx].
\end{align*}
In the case of piecewise constant trial and test functions, the definition of $\dmat{A}_{ij}$ simplifies to $\dmat{A}_{ij} = \int_{\tau_i}\int_{\tau_j}g(\bx, \by)\ds[\by]\ds[\bx]$.

@@

Once the grid, the spaces, and the operator(s) are defined, the main computational step is performed, namely the \textbf{computation of the discrete matrix entries} $\dmat{A}_{ij}$. For pairs of triangles $\tau_i$ and $\tau_j$ that do not share a joint edge or vertex this is done through a simple numerical quadrature rule such as a symmetric Gauss rule for triangles. In the case that two triangles share a joint vertex/edge or the triangles $\tau_i$ and $\tau_j$ are identical (see \cref{fig:triangles}), corresponding singular quadrature rules are used that are based on singularity-removing coordinate transformations \cite{erichsen}.

@@

All of these steps are accelerated through the use of either Numba or OpenCL. In the following section we provide a high-level overview of the library and how these acceleration techniques are deployed before we dive into the design of the computational kernels.

### 2021-05-10 bempp_overview.tex
Revisions done.

The main computational cost involved in solving a problem using boundary element methods is due to discretising the boundary integral operator on the left-hand side of \cref{eq:bnd_integral} to obtain the matrix $\dmat{A}$: using dense methods, discretising an operator has quadratic complexity in terms of the number of surface triangles. For larger problems, this cost can be reduced through approximation techniques, such as hierarchical (H-) matrices or fast multipole methods (FMM) with log-linear or even linear complexity. The price of the improved complexity is significantly more involved data structures and additional approximation errors. Bempp-cl provides interfaces to ExaFMM \cite{bempp_exafmm} for large problems. Here, we focus on dense discretisation that is natively implemented in Bempp-cl and is suitable for medium sized problems up to a few ten thousand elements, depending on available memory. Great care needs to be taken to ensure that the quadratic complexity operator assembly routines perform as highly as possible.

@@

Outside the operator discretization, Numba is used in the following contexts:
\begin{itemize}
  \item Computing the grid topology: this involves iterating through the grid to compute the neighbour relataionships between triangles.
  \item Definition of local-to-global maps for function spaces: again, this requires traversal through the grid and assigning relationships between global and local indices.
  \item Grid functions: a right-hand side function $f$ can be defined as a Python callable. This is just-in-time compiled via Numba and then the product with the corresponding basis functions is integrated in each triangle via numerical quadrature, again via Numba accelerated routines.
  \item Computing sparse matrix entries, such as for mass matrices that are required to translate between representations of grid functions through basis coefficients or projections, or when we want to evaluate operator products.
\end{itemize}

>>>>

The main computational cost involved in solving a problem using boundary element methods is due to discretising the boundary integral operator on the left-hand side of \cref{eq:bnd_integral} to obtain the matrix $\dmat{A}$: using dense methods, discretising an operator has quadratic complexity in terms of the number of surface triangles. For larger problems, this cost can be reduced through approximation techniques, such as hierarchical (H-) matrices or fast multipole methods (FMM) with log-linear or even linear complexity. The price of the improved complexity is significantly more involved data structures and additional approximation errors. Bempp-cl provides interfaces to ExaFMM \cite{bempp_exafmm} for large problems. Here, we focus on dense discretisation that is natively implemented in Bempp-cl and is suitable for medium sized problems up to a few ten thousand elements, depending on available memory. Great care needs to be taken to ensure that the quadratic complexity operator assembly routines perform as efficiently as possible.

@@

Outside the operator discretization, Numba is used in the following contexts:
\begin{itemize}
  \item Computing the grid topology: this involves iterating through the grid to compute the neighbour relationships between triangles.
  \item Definition of local-to-global maps for function spaces: again, this requires traversal through the grid and assigning relationships between global and local indices.
  \item Grid functions: a right-hand side function $f$ can be defined as a Python callable. This is just-in-time compiled via Numba and then the product with the corresponding basis functions is integrated in each triangle via numerical quadrature, again via Numba accelerated routines.
  \item Computing sparse matrix entries, such as for mass matrices that are required to translate between representations of grid functions through basis coefficients or projections, or when we want to evaluate operator products.
\end{itemize}

### 2021-05-10 numba.tex
Revisions done.

With Numba, we use classic loop parallelism: each loop iteration is the assembly of one test element with all trial elements. Within each loop we try to optimize for auto-vectorization by linearly passing through the data in memory order for the individual operations. However, a much smaller fraction of operations is SIMD optimized due to the lack of targeted SIMD constructs in Numba.

>>>>

With Numba, we use loop parallelism: each loop iteration is the assembly of one test triangle with all trial triangles. We then parallelize over the test triangles.
All such assemblies are executed through a parallel for-loop.
Within each loop we try to optimize for auto-vectorization by linearly passing through the data in memory order for the individual operations. However, a much smaller fraction of operations is SIMD optimized due to the lack of targeted SIMD constructs in Numba.

We stress that while Numba provides backends not only for CPU, but also for ROCm and CUDA, we currently only use the CPU component of Numba.

### 2021-05-10 opencl.tex
Revisions done.

OpenCL \cite{opencl} is a heterogeneous compute standard for CPUs, GPUs, FPGAs, and other types of devices that provide conformant drivers. At its core, OpenCL executes compute kernels that can be written in C99, or (more recently) in C++, with some restrictions on the allowed operations. The current version of OpenCL is 3.0, though the most widely implemented standard is OpenCL 1.2, which Bempp-cl uses.

@@

To launch an OpenCL kernel the user must provide relevant data as buffers, which are transferred from host to the corresponding compute device. A kernel string can then be loaded and just-in-time compiled for the device. The kernel is then run, and the results can be copied back to the host.

@@

\begin{itemize}
	\item Test and trial indices of triangles over which to be integrated.
	\item Signs of the normal directions for the spaces.
	\item Test and trial grids as flat floating point arrays, defining each triangle through nine floating point numbers, specifying the $(x, y, z)$ coordinates of each of the three nodes of a triangle.
	\item Test and trial connectivity information that stores the indices of each node for each triangle.
	\item Test and trial mappings of local triangle degrees of freedom to global degrees of freedom.
	\item Test and trial basis function multipliers for each triangle.
	\item Quadrature points and quadrature weights.
	\item A buffer that contains the global assembled matrix.
	\item Additional kernel parameters, such as the wavenumber for Helmholtz problems.
	\item The number of test and trial degrees of freedom.
	\item A single byte that is set to one if the test and trial grids are disjoint.
\end{itemize}

@@

\subsection{Avoiding data races in the global assembly}
Data races in global assembly routines are a problem whenever different triangles need to sum into the same global degree of freedom. To solve this we use standard color coding techniques to split up the computations into chunks that access different data regions. To this end, we define two triangles as neighbors if they share at least one global degree of freedom. Based on this relationship we run a simple greedy coloring algorithm in the initialization phase of a function space. An example coloring is shown in \cref{fig:colouring}.

The compute kernels can then be run color-by-color. OpenCL parallelizes over test elements and trial elements, and so we have to iterate over the product space of possible color combinations. In Numba, we only parallelise over test elements so it is sufficient to iterate over all possible colors in the test space.

>>>>

OpenCL \cite{opencl} is a heterogeneous compute standard for CPUs, GPUs, FPGAs, and other types of devices that provide conformant drivers. At its core, OpenCL executes compute kernels that can be written in OpenCL C, which is based on C99, or (more recently) in C++, with some restrictions on the allowed operations. The current version of OpenCL is 3.0, though the most widely implemented standard is OpenCL 1.2, which Bempp-cl uses.

@@

To launch an OpenCL kernel the user must provide relevant data as buffers, which are transferred from the host to the corresponding compute device. A kernel string can then be loaded and just-in-time compiled for the device. The kernel is then run, and the results can be copied back to the host.

@@

\begin{itemize}
	\item Test and trial indices of triangles over which to be integrated.
	\item Signs of the normal directions for the spaces.
	\item Test and trial grids as flat floating point arrays, defining each triangle through nine floating point numbers, specifying the $(x, y, z)$ coordinates of each of the three nodes of a triangle.
	\item Test and trial connectivity, which are lists of node indices that define the corresponding elements of the test and trial grid.
	\item Test and trial mappings of local triangle degrees of freedom to global degrees of freedom.
	\item Test and trial basis function multipliers for each triangle, which are triangle dependent prefactors needed for certain function spaces (e.g. in electromagnetics).
	\item Quadrature points and quadrature weights.
	\item A buffer that contains the global assembled matrix.
	\item Additional kernel parameters, such as the wavenumber for Helmholtz problems.
	\item The number of test and trial degrees of freedom.
	\item A single byte that is set to one if the test and trial grids are disjoint.
\end{itemize}

@@

\subsection{Avoiding data races in the global assembly}
Data races in global assembly routines are a problem whenever different triangles need to sum into the same global degree of freedom. To solve this we use standard colouring techniques to split up the computations into chunks that access different data regions. To this end, we define two triangles as neighbours if they share at least one global degree of freedom. Based on this relationship we run a simple greedy colouring algorithm in the initialization phase of a function space. An example colouring is shown in \cref{fig:colouring}.

The compute kernels can then be run colour-by-colour. OpenCL parallelizes over test elements and trial elements, and so we have to iterate over the product space of possible colour combinations. In Numba, we only parallelise over test elements so it is sufficient to iterate over all possible colours in the test space.

### 2021-05-10 performance.tex
Revisions done.

\begin{figure}
\begin{center}
\begin{tabular}{l|c|c}
	        &   single      &    double\\
	        \hline
	 PoCL   &   0.05s       &    0.08s\\
	 Numba  &   0.25s       &    0.25s\\
	 GPU    &   0.11s       &    3.00s\\
\end{tabular}
\end{center}
\caption{Comparison of PoCL, Numba and GPU assembly for a grid with 2048 elements.}
\label{fig:cpu_gpu_numba_compare}
\end{figure}
In Bempp-cl, we can easily switch between CPU Assembly, GPU Assembly, and Numba Assembly. \Cref{fig:cpu_gpu_numba_compare} shows a comparison between these modes for a grid with 2048 elements. The speed differences are striking. GPU assembly in single precision is a factor of two slower than CPU assembly (and much slower in double precision due to Nvidia hardware restrictions for double precision). Numba is five times slower for single precision and still around three times slower for double precision than PoCL. The GPU behaviour can be explained by data transfer: while the GPU kernels themselves are extremely fast, data transfer over the bus severely limits performance. Even for medium sized problems, we have to transfer the data back to the main memory as GPU RAM is too limited to keep dense matrices with tens of thousands of rows and columns on the device.

@@

A disadvantage of our approach is that using OpenCL kernels introduce a second language (C99) to the library. Using Numba throughout would give a much more native Python experience, but, while Numba is constantly improving, it is currently difficult to achieve optimal performance for complex operations. OpenCL really shines here, as it makes explicit SIMD operations very easy through dedicated constructs. Moreoever, OpenCL kernels are completely stack based functions, allowing much better compiler optimisations while Numba needs to create every object dynamically, even very small arrays for objects such as coordinate vectors.

Additionally, one has to consider the type of operations that need to be accelerated. For the dense assembly of integral operators, we have very simple data structures that can easily be passed to compute kernels. More complex data structures with larger mixture of data movement operations and computations (e.g., tree-based algorithms), are much harder to accelerate since the Python layer imposes limits here on the performance.

>>>>

\begin{figure}
\begin{center}
\begin{tabular}{l|c|c}
	        &   single      &    double\\
	        \hline
	 PoCL   &   0.05s       &    0.08s\\
	 Numba  &   0.25s       &    0.25s\\
	 GPU    &   0.11s       &    3.00s\\
\end{tabular}
\end{center}
\caption{Comparison of PoCL, Numba and GPU assembly for a grid with 2048 elements.}
\label{fig:cpu_gpu_numba_compare}
\end{figure}
In Bempp-cl, we can easily switch between CPU Assembly, GPU Assembly, and Numba Assembly. \Cref{fig:cpu_gpu_numba_compare} shows a comparison between these modes for a grid with 2048 elements. The speed differences are striking. GPU assembly in single precision is a factor of two slower than CPU assembly, and much slower in double precision due to the limited double precision performance of our hardware, though we note that this is not an issue for data center accelerators by Nvidia that are also optimised for double precision. Numba is five times slower for single precision and still around three times slower for double precision than PoCL. The GPU behaviour can be explained by data transfer: while the GPU kernels themselves are extremely fast, data transfer over the bus severely limits performance. Even for medium sized problems, we have to transfer the data back to the main memory as GPU RAM is too limited to keep dense matrices with tens of thousands of rows and columns on the device.

@@

A disadvantage of our approach is that using OpenCL kernels introduces a second language (C99) to the library. Using Numba throughout would give a much more native Python experience, but, while Numba is constantly improving, it is currently difficult to achieve optimal performance for complex operations. OpenCL really shines here, as it makes explicit SIMD operations very easy through dedicated constructs. Moreoever, OpenCL kernels are completely stack/register based functions, allowing much better compiler optimisations while Numba needs to create every object dynamically, even for very small arrays for objects such as coordinate vectors. We need to stress that we have performed very few optimisations specific to Numba, while significant optimisation has gone into the OpenCL codes. It is therefore well possible that the performance gap between Numba and OpenCL can be significantly reduced. But from other projects our own anecdotal experience is that the more Numba is optimised, the less Pythonic and more C-like Numba functions look. So while Numba is a very powerful tool, it requires its own techniques for optimisation, different from standard Python code.

Another important consideration with respect to Python and just-in-time acceleration is the type of algorithms that benefit. For the dense assembly of integral operators, we have very simple data structures that can easily be passed to compute kernels. More complex data structures with larger mixture of data movement operations and computations (e.g., tree-based algorithms), are much harder to accelerate since the Python layer imposes limits here on the performance.
